---
title: "Approximating visibility"
author: "Dennis M. Feehan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Approximating visibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r, message = FALSE}
library(networkreporting)
library(dplyr)
library(tibble)
```

## What visibility is, and why it usually cannot be calculated

A network reporting estimator divides each report by the number of frame-population
members who could have made it. That number is the alter's **visibility**, and it
is what converts "how many reports did we see" into "how big is the population".

For siblings, visibility is not a modelling choice. It can be read straight off
ego's own roster:

* an alter who is **on the frame** has visibility `y.F`
* an alter who is **not** has visibility `y.F + 1`

where `y.F` is the number of ego's siblings who are in the frame population. That
is `vis_from_clique()`, and it is the default everywhere.

```{r}
vis_from_clique()
```

The three assumptions it prints are the whole reason it works. Siblingship is an
equivalence relation, so it cuts the population into disjoint groups; ego belongs
to the group ego reports about; and reporting within the group is complete. Given
those three, ego's roster is enough to recover the visibility of *every* alter,
including alters whose own neighbourhoods ego cannot see.

Households satisfy all three. Most other ties do not:

| Tie | Partitions? | Ego in group? | Verdict |
|---|---|---|---|
| siblings | yes | yes | exact |
| household | yes | yes | exact |
| cousins | **no** (not transitive) | yes | not identified |
| parents | yes | **no** | needs another roster |
| neighbours | **no** | yes | not identified |

For those, visibility generally is not identified from one-sided reports at all.
The standard move is to substitute a summary of the *respondents'* visibility.
That is a real assumption with a knowable bias, and the point of this machinery is
to give it somewhere to live other than an undocumented `case_when` in analysis
code.

## The subtlety that determines the whole design

It is tempting to think that a visibility which is constant within an estimation
cell cancels out of a rate — the rate is a ratio, and both numerator and
denominator get divided by the same number.

**It does not**, and seeing why is what makes the rest of this make sense.

Deaths and exposure do not receive the same visibility:

* a dead alter is never on the frame, so **every death** gets `1/(y.F + 1)`
* exposure is a **mixture**: living on-frame alters get `1/y.F`, living off-frame
  alters get `1/(y.F + 1)`

So visibility survives into the rate purely through that asymmetry. Here is the
consequence, made explicit:

```{r}
alters <- tibble(
  .ego.id   = c(1, 1, 1, 2, 2),
  .sib.in.F = c(1,  1,  0,  1,  0),
  y.F       = c(2,  2,  2,  1,  1),
  sib.occ   = c(0,  0,  1,  0,  1),   # deaths, only ever off-frame
  sib.exp   = c(10, 10, 4,  10, 6)    # exposure, a mixture
)

rate <- function(w) sum(alters$sib.occ * w) / sum(alters$sib.exp * w)

# a rule that ignores frame status: one number for every alter
flat <- rep(1/4, nrow(alters))

# the clique rule, which preserves the split
clique <- vis_from_clique()
split  <- clique$predict(alters, clique$fit(NULL, NULL))$vis_weight

c(flat      = rate(flat),
  aggregate = sum(alters$sib.occ) / sum(alters$sib.exp),
  split     = rate(split))
```

The flat rule reproduces the **aggregate** estimator exactly, because a constant
divides out of a ratio. An approximation that assigns one number per cell without
keeping the on-frame/off-frame split therefore does nothing at all.

This is why `vis_from_donor()` estimates a *group size* and then reconstructs the
split from it, rather than estimating a visibility directly.

## Borrowing visibility from donors

`vis_from_donor()` takes a donor population — normally the survey respondents,
whose own visibility *is* derivable — summarises it, and applies the result to
alters that cannot be resolved exactly.

```{r}
rule <- vis_from_donor(match_on = c(.sib.sex = "sex"), min_donors = 10)
rule
```

Two things to notice in that printout.

`is_estimated` is `TRUE`. The estimated group size is a **sample** quantity, so
holding it fixed across bootstrap replicates would understate the variance.
`is_estimated` is what drives the replicate loop to refit rather than freeze; for
`vis_from_clique()` it is `FALSE`, and freezing is then correct, because
visibility is a function of ego's own reports rather than of who was sampled.

And the assumptions include the direction of the bias, not just its existence.
Donors are alive and on the frame. Many of the alters needing an imputed
visibility are dead. Wherever visibility correlates with mortality — through
family size for kin ties, living arrangements for household ties — the donor is
systematically wrong, not merely noisy.

### Harmonic, not arithmetic

The default summary is the weighted **harmonic** mean, and that is not a detail.
The individual estimator averages `1/v`, so the functional that makes the plug-in
unbiased is `(E[1/v])^-1`, not `E[v]`.

```{r}
# four donors with very different sibship sizes
spread <- tibble(.ego.id = 1:4, y.F = c(1, 2, 3, 11), w = 1)
one    <- tibble(.ego.id = 1, .sib.in.F = 0)

S_under <- function(stat) {
  r <- vis_from_donor(match_on = NULL, statistic = stat, min_donors = 1)
  r$predict(one, r$fit(spread, "w"))$vis
}

c(harmonic   = S_under("harmonic"),
  arithmetic = S_under("arithmetic"))
```

By Jensen's inequality harmonic `<=` arithmetic always, so the two disagree in a
*known direction*, by an amount that grows with the variance of visibility. They
coincide exactly when there is no variance at all.

`"arithmetic"` remains available, because it is what the historical
`y.F.bar / (y.F.bar + 1)` adjustment factor used. If you are reproducing an older
analysis, that is the setting that will match it.

## Donor coverage fails, routinely

`match_on` describes the **alter**. Donors are **respondents**. DHS and MICS
interview women aged 15–49, so an alter aged 60, or any male alter, has no donor
cell at all.

Left alone, that produces `NA` and the `NA` propagates silently into rates.
`min_donors` together with `on_missing` makes it loud instead:

```{r, error = TRUE}
donors <- tibble(.ego.id = 1:40, y.F = 3, sex = "f", w = 1)
alters <- tibble(.ego.id = c(1, 2), .sib.in.F = c(0, 0), .sib.sex = c("f", "m"))

strict <- vis_from_donor(match_on = c(.sib.sex = "sex"), min_donors = 10)
strict$predict(alters, strict$fit(donors, "w"))
```

The three options are `"error"` (the default, above), `"fallback"` (use the global
value and say so), and `"na"` (leave it unresolved, so that a coalesced rule can
try the next tier).

## You must say what the tie is

`vis_from_clique()` will not run until you declare what kind of tie the reports
are about:

```{r, error = TRUE}
apply_visibility_rule(vis_from_clique(), esc.dat = tibble(
  .ego.id = 1, .sib.in.F = 0, y.F = 2))
```

This is deliberate, and it is the most important safety property in this
vignette. The clique rule returns a finite, plausible number for *any* roster.
Nothing in the data distinguishes a tie that satisfies its assumptions from one
that does not, so applied in the wrong place it is wrong with no outward sign --
and the provenance table then reports `clique: 100%`, which reads as "the exact
rule was used".

How wrong? Socsim gives the true reporting network, and therefore each alter's
true visibility, so this is measurable rather than a matter of opinion:

| roster | off-frame (carries the deaths) | on-frame (exposure only) | differential |
|---|---|---|---|
| maternal siblings | exact (100%) | exact (100%) | 1.000 |
| maternal cousins | exact (100%) | exact (100%) | 1.000 |
| paternal cousins | exact (100%) | exact (100%) | 1.000 |
| maternal **union** paternal cousins | 1.089x too high (65% exact) | 1.061x too high (63% exact) | **1.026** |

The interesting row is the last one, and the interesting non-rows are the two
above it.

"Cousinship is not transitive" is the usual reason given for cousins not being a
clique, and in general it is true: your maternal cousin and your paternal cousin
are not each other's cousins. But *within one line* it is transitive --- everyone
sharing a maternal grandmother forms an equivalence class --- so a
maternal-cousin roster is a clique, and the rule is exact on it. What breaks is
pooling the two lines.

So the lesson is not that cousins are a special case to be careful of. It is that
whether a given roster is a clique is a question about **how that roster was
built**, which no amount of looking at the data will answer. Hence the
declaration.

Note *where* the pooled roster goes wrong: more on the off-frame side, which
carries every death, than on the on-frame side. By the argument above, a uniform
error would cancel out of the rate; this differential one does not.

So declare the tie:

```{r}
tie_config("clique", name = "siblings")
tie_config("group",  name = "maternal cousins")
```

`"clique"` is siblings and household members; `"group"` is cousins; `"star"` is
parents, whose visibility is a fact about the *sibship* rather than the parent
roster; `"unbounded"` is neighbours and acquaintances. A rule that makes no
structural assumption, such as `vis_from_donor()`, needs no declaration and
accepts any.

### What else belongs on the tie

Two further things are facts about the tie rather than about the rule, and a
`tie_config()` can carry both:

```{r}
tie_config("star", name = "parents",
           ego.in.group    = FALSE,     # ego is not in the group being sized
           frame.indicator = "in.frame") # this tie's own eligibility column
```

`ego.in.group` is the `+ 1`: it says whether ego is counted among the group
members who could report an alter. `frame.indicator` names the column saying who
is in the frame population *for this tie*, which starts to matter once ties
differ in who is eligible — neighbours bounded by a compound, say, against
siblings who are not.

Both default to undeclared, in which case nothing changes: the rule's own
setting is used, and the frame column defaults to `.sib.in.F`.

Where a tie declares one of these and the rule or the call also sets it,
**agreement is fine and disagreement is an error naming both sources**:

```{r, error = TRUE}
apply_visibility_rule(
  vis_from_clique(ego.in.group = TRUE),
  esc.dat = tibble(.ego.id = 1, .sib.in.F = 0, y.F = 2),
  tie     = tie_config("clique", name = "siblings", ego.in.group = FALSE))
```

There is deliberately no rule for which one wins. Silent precedence would
compute a number under an assumption you did not choose — the same failure the
tie gate itself exists to prevent — so the package makes you say it once.

Note that what a tie declares reaches the provenance table, including the
assumptions, which are recomputed from the setting actually used rather than
from whatever the rule was built with:

```{r}
apply_visibility_rule(
  vis_from_clique(),
  esc.dat = tibble(.ego.id = c(1, 1), .sib.in.F = c(1, 0), y.F = c(2, 2)),
  tie     = tie_config("clique", name = "households",
                       ego.in.group = FALSE))$provenance
```

## When the group is not the roster

`vis_from_clique()` works out the group by counting ego's roster. Sometimes the
group whose size sets an alter's visibility is *not* the roster the alter was
reported on, and only you can say what it is:

* **Pooled ties.** Cousins are reported through the maternal and paternal sides
  separately, but an alter's visibility depends on everyone who could have
  reported them — both sides, plus ego's own siblings. That group is not any
  single roster.
* **Nested ties.** "Pooled cousins, excluding ego's siblings" is a difference of
  two counts. No roster holds it.
* **Borrowed groups.** A parent's visibility is the number of their children on
  the frame — a fact about the *sibship*, not the parent roster.

`vis_from_group_size()` takes the count from a column instead of deriving it:

```{r}
cousins <- tibble(
  .ego.id   = c(1, 1, 2),
  .sib.in.F = c(1, 0, 1),
  n_pooled_and_F  = c(5, 5, 3),   # ego's pooled cousinship, on frame
  n_sibship_and_F = c(2, 2, 1))
cousins$n_pooled_nosib_and_F <-
  cousins$n_pooled_and_F - cousins$n_sibship_and_F

pooled <- vis_from_group_size("n_pooled_and_F", label = "pooled cousins")
pooled$predict(cousins, pooled$fit(NULL, NULL))$vis
```

The rule still does the frame split — an alter who is themselves in the frame
population cannot report themselves — because that asymmetry is the only route
by which visibility reaches a rate. A basis that deliberately omits it says so:

```{r}
nosib <- vis_from_group_size("n_pooled_nosib_and_F",
                             subtract.self = FALSE,
                             label = "pooled without siblings")
nosib$predict(cousins, nosib$fit(NULL, NULL))$vis
```

Two things to be clear about. This rule makes **no structural assumption**, so
it needs no `tie` and accepts any — but that is not because it is safe the way
`vis_from_donor()` is safe. It is because you have taken on the part
`vis_from_clique()` would have derived. The package checks the arithmetic; it
cannot check that your column counts the right people, and the assumptions say
so:

```{r}
pooled$assumptions
```

And `label` earns its place when you are comparing bases rather than committing
to one: run the same cells under each, and the provenance says which produced
which.

## Coalescing: exact where possible, approximate where not

`vis_coalesce()` takes rules in priority order. For each row the first rule
returning a non-`NA` visibility wins, and the row records which tier resolved it.

```{r}
mixed <- tibble(
  .ego.id   = c(1, 2, 3),
  .sib.in.F = c(0, 0, 0),
  .sib.sex  = c("f", "f", "f"),
  y.F       = c(2, NA, NA)     # only ego 1 has an exact roster
)

chain <- vis_coalesce(
  vis_from_clique(),                                    # exact where it exists
  vis_from_donor(match_on = c(.sib.sex = "sex"),        # matched approximation
                 min_donors = 10, on_missing = "na"),
  vis_from_donor(match_on = NULL, min_donors = 1)       # global, last resort
)

out <- chain$predict(mixed, chain$fit(donors, "w"))
out
```

Ego 1 is resolved exactly by tier 1; the others fall through to the donor tier.
`vis_tier` records which one caught each row.

That example falls through because tier 1 had no `y.F` to work with. The other
way a chain falls through is by *declaration*: on a tie the clique rule does not
apply to, the tier is dropped outright.

```{r}
res <- apply_visibility_rule(
  vis_coalesce(vis_from_clique(), vis_from_donor(match_on = NULL, min_donors = 1)),
  esc.dat = tibble(.ego.id = c(1, 2), .sib.in.F = c(0, 1), y.F = c(2, 2)),
  ego.dat = donors, weights = "w",
  tie     = tie_config("group", name = "cousins"))

res$provenance$dropped_tiers
```

This is the distinction worth holding onto: the clique tier is removed because
it is *inapplicable*, not skipped because it returned `NA`. It never returns
`NA` -- which is why, before the tie had to be declared, a chain like this one
claimed every cousin alter for the exact tier.

## Reading the provenance table

`apply_visibility_rule()` returns the values alongside a provenance object, and
`sibling_estimator()` attaches that object to its result. It is what turns "some
of this was approximated" into a number.

```{r}
res <- apply_visibility_rule(vis_from_clique(), esc.dat = tibble(
  .ego.id   = c(1, 1, 1, 2, 2),
  .sib.in.F = c(1, 1, 0, 1, 0),
  y.F       = c(2, 2, 2, 1, 1),
  sib.occ   = c(0, 0, 1, 0, 1),
  sib.exp   = c(10, 10, 4, 10, 6)
), tie = tie_config("clique", name = "siblings"))

res$provenance
```

The two share-of lines are separate on purpose. The share of *deaths* that were
approximated and the share of *exposure* that were approximated are different
numbers, and an approximation touching very little exposure but most of the
deaths is not a small approximation. Reporting only one of them would hide
exactly the case worth worrying about.

## Where this is going

`vis_from_clique()` and `vis_from_donor()` are two implementations of one
interface, split into `fit` and `predict`. That split is deliberate: a fitted
model is just a `fit()` that returns a model object and a `predict()` that calls
`predict()` on it. Two consequences are already honoured here so that a model
rule needs no interface change — a non-integer visibility works end to end, and
`is_estimated` already drives the bootstrap path rather than waiting for a model
to appear.
