---
title: "Approximating visibility"
author: "Dennis M. Feehan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Approximating visibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r, message = FALSE}
library(networkreporting)
library(dplyr)
library(tibble)
```

## What visibility is, and why it usually cannot be calculated

A network reporting estimator divides each report by the number of frame-population
members who could have made it. That number is the alter's **visibility**, and it
is what converts "how many reports did we see" into "how big is the population".

For siblings, visibility is not a modelling choice. It can be read straight off
ego's own roster:

* an alter who is **on the frame** has visibility `y.F`
* an alter who is **not** has visibility `y.F + 1`

where `y.F` is the number of ego's siblings who are in the frame population. That
is `vis_from_clique()`, and it is the default everywhere.

```{r}
vis_from_clique()
```

The three assumptions it prints are the whole reason it works. Siblingship is an
equivalence relation, so it cuts the population into disjoint groups; ego belongs
to the group ego reports about; and reporting within the group is complete. Given
those three, ego's roster is enough to recover the visibility of *every* alter,
including alters whose own neighbourhoods ego cannot see.

Households satisfy all three. Most other ties do not:

| Tie | Partitions? | Ego in group? | Verdict |
|---|---|---|---|
| siblings | yes | yes | exact |
| household | yes | yes | exact |
| cousins | **no** (not transitive) | yes | not identified |
| parents | yes | **no** | needs another roster |
| neighbours | **no** | yes | not identified |

For those, visibility generally is not identified from one-sided reports at all.
The standard move is to substitute a summary of the *respondents'* visibility.
That is a real assumption with a knowable bias, and the point of this machinery is
to give it somewhere to live other than an undocumented `case_when` in analysis
code.

## The subtlety that determines the whole design

It is tempting to think that a visibility which is constant within an estimation
cell cancels out of a rate — the rate is a ratio, and both numerator and
denominator get divided by the same number.

**It does not**, and seeing why is what makes the rest of this make sense.

Deaths and exposure do not receive the same visibility:

* a dead alter is never on the frame, so **every death** gets `1/(y.F + 1)`
* exposure is a **mixture**: living on-frame alters get `1/y.F`, living off-frame
  alters get `1/(y.F + 1)`

So visibility survives into the rate purely through that asymmetry. Here is the
consequence, made explicit:

```{r}
alters <- tibble(
  .ego.id   = c(1, 1, 1, 2, 2),
  .sib.in.F = c(1,  1,  0,  1,  0),
  y.F       = c(2,  2,  2,  1,  1),
  sib.occ   = c(0,  0,  1,  0,  1),   # deaths, only ever off-frame
  sib.exp   = c(10, 10, 4,  10, 6)    # exposure, a mixture
)

rate <- function(w) sum(alters$sib.occ * w) / sum(alters$sib.exp * w)

# a rule that ignores frame status: one number for every alter
flat <- rep(1/4, nrow(alters))

# the clique rule, which preserves the split
clique <- vis_from_clique()
split  <- clique$predict(alters, clique$fit(NULL, NULL))$vis_weight

c(flat      = rate(flat),
  aggregate = sum(alters$sib.occ) / sum(alters$sib.exp),
  split     = rate(split))
```

The flat rule reproduces the **aggregate** estimator exactly, because a constant
divides out of a ratio. An approximation that assigns one number per cell without
keeping the on-frame/off-frame split therefore does nothing at all.

This is why `vis_from_donor()` estimates a *group size* and then reconstructs the
split from it, rather than estimating a visibility directly.

## Borrowing visibility from donors

`vis_from_donor()` takes a donor population — normally the survey respondents,
whose own visibility *is* derivable — summarises it, and applies the result to
alters that cannot be resolved exactly.

```{r}
rule <- vis_from_donor(match_on = c(.sib.sex = "sex"), min_donors = 10)
rule
```

Two things to notice in that printout.

`is_estimated` is `TRUE`. The estimated group size is a **sample** quantity, so
holding it fixed across bootstrap replicates would understate the variance.
`is_estimated` is what drives the replicate loop to refit rather than freeze; for
`vis_from_clique()` it is `FALSE`, and freezing is then correct, because
visibility is a function of ego's own reports rather than of who was sampled.

And the assumptions include the direction of the bias, not just its existence.
Donors are alive and on the frame. Many of the alters needing an imputed
visibility are dead. Wherever visibility correlates with mortality — through
family size for kin ties, living arrangements for household ties — the donor is
systematically wrong, not merely noisy.

### Harmonic, not arithmetic

The default summary is the weighted **harmonic** mean, and that is not a detail.
The individual estimator averages `1/v`, so the functional that makes the plug-in
unbiased is `(E[1/v])^-1`, not `E[v]`.

```{r}
# four donors with very different sibship sizes
spread <- tibble(.ego.id = 1:4, y.F = c(1, 2, 3, 11), w = 1)
one    <- tibble(.ego.id = 1, .sib.in.F = 0)

S_under <- function(stat) {
  r <- vis_from_donor(match_on = NULL, statistic = stat, min_donors = 1)
  r$predict(one, r$fit(spread, "w"))$vis
}

c(harmonic   = S_under("harmonic"),
  arithmetic = S_under("arithmetic"))
```

By Jensen's inequality harmonic `<=` arithmetic always, so the two disagree in a
*known direction*, by an amount that grows with the variance of visibility. They
coincide exactly when there is no variance at all.

`"arithmetic"` remains available, because it is what the historical
`y.F.bar / (y.F.bar + 1)` adjustment factor used. If you are reproducing an older
analysis, that is the setting that will match it.

## Donor coverage fails, routinely

`match_on` describes the **alter**. Donors are **respondents**. DHS and MICS
interview women aged 15–49, so an alter aged 60, or any male alter, has no donor
cell at all.

Left alone, that produces `NA` and the `NA` propagates silently into rates.
`min_donors` together with `on_missing` makes it loud instead:

```{r, error = TRUE}
donors <- tibble(.ego.id = 1:40, y.F = 3, sex = "f", w = 1)
alters <- tibble(.ego.id = c(1, 2), .sib.in.F = c(0, 0), .sib.sex = c("f", "m"))

strict <- vis_from_donor(match_on = c(.sib.sex = "sex"), min_donors = 10)
strict$predict(alters, strict$fit(donors, "w"))
```

The three options are `"error"` (the default, above), `"fallback"` (use the global
value and say so), and `"na"` (leave it unresolved, so that a coalesced rule can
try the next tier).

## Coalescing: exact where possible, approximate where not

`vis_coalesce()` takes rules in priority order. For each row the first rule
returning a non-`NA` visibility wins, and the row records which tier resolved it.

```{r}
mixed <- tibble(
  .ego.id   = c(1, 2, 3),
  .sib.in.F = c(0, 0, 0),
  .sib.sex  = c("f", "f", "f"),
  y.F       = c(2, NA, NA)     # only ego 1 has an exact roster
)

chain <- vis_coalesce(
  vis_from_clique(),                                    # exact where it exists
  vis_from_donor(match_on = c(.sib.sex = "sex"),        # matched approximation
                 min_donors = 10, on_missing = "na"),
  vis_from_donor(match_on = NULL, min_donors = 1)       # global, last resort
)

out <- chain$predict(mixed, chain$fit(donors, "w"))
out
```

Ego 1 is resolved exactly by tier 1; the others fall through to the donor tier.
`vis_tier` records which one caught each row.

## Reading the provenance table

`apply_visibility_rule()` returns the values alongside a provenance object, and
`sibling_estimator()` attaches that object to its result. It is what turns "some
of this was approximated" into a number.

```{r}
res <- apply_visibility_rule(vis_from_clique(), esc.dat = tibble(
  .ego.id   = c(1, 1, 1, 2, 2),
  .sib.in.F = c(1, 1, 0, 1, 0),
  y.F       = c(2, 2, 2, 1, 1),
  sib.occ   = c(0, 0, 1, 0, 1),
  sib.exp   = c(10, 10, 4, 10, 6)
))

res$provenance
```

The two share-of lines are separate on purpose. The share of *deaths* that were
approximated and the share of *exposure* that were approximated are different
numbers, and an approximation touching very little exposure but most of the
deaths is not a small approximation. Reporting only one of them would hide
exactly the case worth worrying about.

## Where this is going

`vis_from_clique()` and `vis_from_donor()` are two implementations of one
interface, split into `fit` and `predict`. That split is deliberate: a fitted
model is just a `fit()` that returns a model object and a `predict()` that calls
`predict()` on it. Two consequences are already honoured here so that a model
rule needs no interface change — a non-integer visibility works end to end, and
`is_estimated` already drives the bootstrap path rather than waiting for a model
to appear.
