<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{respondent-driven sampling}
-->
Analyzing respondent-driven sampling (RDS) data using the **networkreporting** package
======================================================================================

Introduction
------------

The **networkreporting** package has tools for analyzing data that were collected
using respondent-driven sampling (RDS).

This introduction will assume that you already have the **networkreporting**
package installed. If you don't, please refer to the introductory vignette 
("getting started") for instructions on how to do this. 

Review of respondent driven sampling
------------------------------------

TODO - eventually, this will be a very brief review of RDS, just enough to
explain the two bootstrap options as they get used with generalized scale-up

In the descriptions below, we will need some notation to refer to the waves
at each stage

TODO FINISH...

Preparing data
--------------

```{r, message=FALSE}
library(networkreporting)
library(plyr)
library(ggplot2) # we'll use qplot from ggplot2 for plots
theme_set(theme_minimal())

data.dir <- file.path("~", "Dropbox", "dennis_matt", "data", "processed")

# eventually, a sample RDS dataset will be included with the
# package. for now, this will only work on a few machines
# and only after gnsum-curitiba-goc-prep.r has been run
load(file.path(data.dir, "gnsum-curitiba-goc-data.RData"))
```

We obtain a sample of the hidden population using respondent-driven sampling
by starting with a set of *seeds* and asking each of them to refer us to
hidden population members she is connected to. The process then repeats,
wave after wave, until we are finished.  The data that result can be thought
of as a tree, also called a chain, starting from each seed respondent.

Although we usually think of respondent-driven sampling data as being composed
of these chains, the dataset itself typically arrives as a rectangular array.
The **make.chain** function allows us to take one of these rectangular arrays and
turn it into a linked chain structure. (If you are not familiar with linked data
structures, it suffices to understand that the rest of the respondent-driven sampling
functions we will want to use require the data to have been formatted in this way.)

The main requirement of **make.chain** is that the rows of the dataset have ids
that allow us to determine whether or not one row is the parent of another. You
can pass in your own function which, given two ids, returns **TRUE** if the second id
is the parent of the first and **FALSE** otherwise. The default is to assume that
we are using the id scheme from the Curitiba study [TODO CITE].

TODO -- more detail on how ids work, etc; need to be able to parse the ids in a way
        that describes the tree...
TODO -- chain summary function?

```{r}
## build the nomination chains based on the
## ids in the data
seed.ids <- c("1", "2", "3", "4", "5")
chains <- llply(seed.ids,
                ## NB: make.chain is not currently exported, so we
                ## have to specify the package namespace explicitly
                networkreporting:::make.chain,
                survey.data)
```

TODO - describe the trait we want to measure and the mixing setup


Variance estimates from bootstrap of Salganik [TODO CITE]
---------------------------------------------------------

The bootstrap of Salganik [TODO CITE] accounts for the non-independent nature
of RDS observations by characterizing each respondent in terms of a small number
of categorical states. For example, we might make a note of whether each respondent
is male or female and whether or not each respondent has a large or small number of
network connections. Essentially, this models the referral process as a Markov Chain,
where the states given by the respondents' traits and transitions given by one 
respondent referring another, mod the respondents' traits. [TODO EXPLAIN MUCH BETTER;
CITE DESCRIPTION IN WEIR ET AL]

The next step, then, is to specify the traits that we think are most salient in
driving the referral process, and to then estimate the mixing model based on
those traits:

```{r}
## pick a set of traits and estimate the mixing
## info for them
these.traits <- c("netsize.bss.big")

mm <- networkreporting:::estimate.mixing(nonseed.survey.data, parent.data, these.traits)
```

We also need the empirical distribution of respondent degrees for each trait in
the data. We can compute this with **estimate.degree.distns**. Note that we
specify that we want to retain **total.alters** and **total.aware** from the
original data; we will need these to compute our estimates once we have the
bootstrap resamples.

```{r}
dd <- networkreporting:::estimate.degree.distns(survey.data,
                                                d.hat.vals="netsize.bss",
                                                traits=these.traits,
                                                keep.vars=c("total.alters",
                                                            "total.aware"))
```

Finally, we take the actual bootstrap resamples using **rds.mc.boot.draws**. 
```{r}
num.reps <- 100
boot.mc.dat <- networkreporting:::rds.mc.boot.draws(chains,
                                                    mm,
                                                    dd,
                                                    num.reps=num.reps)
```

Now that we have the bootstrap resamples, we can produce an estimate for each
one.

```{r}
## use the RDS-II estimator on each bootstrap resample

## get estimates of the total number of alters and the total number of
## alters aware that the ego is a heavy drug user for each bootstrap rep
boot.mc.total.ests <- ldply(boot.mc.dat,
                            rdsII.estimator,
                            ## now we use the name of the resampled degree
                            ## (degree) instead of the original one
                            d.hat.vals="degree",
                            y.vals="total.alters",
                            missing="complete.obs")

boot.mc.aware.ests <- ldply(boot.mc.dat,
                            rdsII.estimator,
                            d.hat.vals="degree",
                            y.vals="total.aware",
                            missing="complete.obs")

boot.mc.ests <- data.frame(estimate=boot.mc.aware.ests[,1] / boot.mc.total.ests[,1],
                           estimator="mc")

```

Variance estimates form bootstrap of Weir et al [TODO CITE]
-----------------------------------------------------------

Alternatively, we can obtain bootstrap resamples using the method
of Weir et al [TODO CITE]. In this method, instead of randomly
choosing a seed for each bootstrap resample, the original seeds used
in the actual data collection are considered fixed. 

The structure of each chain in the original data is preserved, but
within each bootstrap resample, the traits of the referrals are
drawn from a Markov Model in the same way as they were above. For example,
suppose we have two traits, 0 and 1. The steps are:


We can obtain bootstrap resamples using thie technique by calling 
**rds.chain.boot.draws**.

```{r}
boot.chain.dat <- networkreporting:::rds.chain.boot.draws(chains,
                                                          mm,
                                                          dd,
                                                          num.reps=num.reps)
```

And we next produce an estimate for each one.

```{r}
## use the RDS-II estimator on each bootstrap resample

## get estimates of the total number of alters and the total number of
## alters aware that the ego is a heavy drug user for each bootstrap rep
boot.chain.total.ests <- ldply(boot.chain.dat,
                               rdsII.estimator,
                               ## now we use the name of the resampled degree
                               ## (degree) instead of the original one
                               d.hat.vals="degree",
                               y.vals="total.alters",
                               missing="complete.obs")

boot.chain.aware.ests <- ldply(boot.chain.dat,
                               rdsII.estimator,
                               d.hat.vals="degree",
                               y.vals="total.aware",
                               missing="complete.obs")

boot.chain.ests <- data.frame(estimate=boot.chain.aware.ests[,1] / boot.chain.total.ests[,1],
                              estimator="chain")

```

Now we can compare the estimated visibility using the two bootstraps.

```{r}
both.ests <- rbind(boot.chain.ests, boot.mc.ests)

comp.plot <- ggplot(both.ests) +
             geom_density(aes(x=estimate, color=estimator, group=estimator))
```




